## Strain means with separate sexes

The following plots show LOD traces for the traits
mapped by strain mean separately in the two sexes.

```{r means_by_sex}

m.mean.pheno <- t(sapply(1:length(K.idx), 
    function(x) colMeans(norm.cds[intersect(which(mapping.covar[,1] == names(K.idx)[x]), which(sex == 1)),,drop=FALSE], na.rm = TRUE)))
rownames(m.mean.pheno) <- names(K.idx)

f.mean.pheno <- t(sapply(1:length(K.idx), 
    function(x) colMeans(norm.cds[intersect(which(mapping.covar[,1] == names(K.idx)[x]), which(sex == 0)),,drop=FALSE])))
rownames(f.mean.pheno) <- names(K.idx)

mean.overall.K <- calc_kinship(mean.genoprobs)

m.mean.file <- here("Results", "scan1_results_males_means.RDS")
f.mean.file <- here("Results", "scan1_results_females_means.RDS")
if(!file.exists(m.mean.file) || overwrite_previous){
    
    m.mean.scans <- scan1(mean.genoprobs, m.mean.pheno, kinship = mean.K)
    saveRDS(m.mean.scans, m.mean.file)

    f.mean.scans <- scan1(mean.genoprobs, f.mean.pheno, kinship = mean.K)
    saveRDS(f.mean.scans, f.mean.file)

}else{
    m.mean.scans <- readRDS(m.mean.file)
    f.mean.scans <- readRDS(f.mean.file)
}
```

### Male Strain Means

```{r male_mean_scans, fig.width = 9, fig.height = 4}
for(i in 1:ncol(mapping.pheno)){
    plot(m.mean.scans, lodcol = i, map = map, main = paste("Male", colnames(mapping.pheno)[i]))
}
```

### Female Strain Means

```{r female_mean_scans, fig.width = 9, fig.height = 4}
for(i in 1:ncol(mapping.pheno)){
    plot(f.mean.scans, lodcol = i, map = map, main = paste("Female", colnames(mapping.pheno)[i]))
}
```

## Map individuals, sex as interactive covariate

The traces below show results for mapping with individuals
again, but this time using sex as an interactive covariate.
In the plots below, the mapping with sex as an additive 
covariate is in red, and the mapping with sex as an interactive
covariate is in blue. 

I'm  not sure why the mapping looks so strange. It looks as
if there are sex interactions all over the place, but I don't
think this is right. I will have to look into this more.

```{r map_sex_interactive}
int.scan.file <- here("Results", "scan1_results_interactive.RDS")
if(!file.exists(int.scan.file) || overwrite_previous){
    int.pheno.scans <- scan1(expanded.genoprobs, mapping.pheno, kinship = exp.K, 
        addcovar = sex, intcovar = sex)
    saveRDS(int.pheno.scans, int.scan.file)

}else{
    int.pheno.scans <- readRDS(int.scan.file)
}

```


```{r int_scans, fig.width = 9, fig.height = 4}
for(i in 1:ncol(mapping.pheno)){
    plot(int.pheno.scans, lodcol = i, map = map, 
        main = colnames(mapping.pheno)[i])
    plot(pheno.scans, lodcol = i, map = map, add = TRUE, col = "red")
}
```



## Map with strain means

We then mapped the traits again using strain means. 
As we can see from the scatter plots of individuals
in the section "Phenotype Distributions", strain 
means are not neceesarily a good description of 
the distribution across individuals in the strain, 
since each strain has such wide variation. 

Mapping with strain means shows better defined QTL
than mapping with individuals, but it is possible 
these could be false positives because we've falsely
reduced the variation across individual genotypes.

The first set of plots show the LOD traces for each
trait adjusted by sex.

```{r strain_avg}
adj.pheno <- adjust(mapping.pheno, sex)
mean.pheno <- t(sapply(1:length(K.idx), 
    function(x) colMeans(adj.pheno[which(mapping.covar[,1] == names(K.idx)[x]),])))
rownames(mean.pheno) <- names(K.idx)

just.strain <- sapply(strsplit(rownames(mapping.genoprobs[[1]]), "/", fixed = TRUE), function(x) x[1])
mean.genoprobs <- mapping.genoprobs
mean.K <- mapping.K
for(i in 1:length(mean.genoprobs)){
    rownames(mean.genoprobs[[i]]) <- just.strain
    rownames(mean.K[[i]]) <- colnames(mean.K[[i]]) <- just.strain
}

mean.scan.file <- here("Results", "scan_results_means.RDS")

if(!file.exists(mean.scan.file) || overwrite_previous){
    mean.scans <- scan1(mean.genoprobs, mean.pheno, kinship = mean.K)
    saveRDS(mean.scans, mean.scan.file)
}else{
    mean.scans <- readRDS(mean.scan.file)
}

```

```{r plot_mean, fig.width = 9, fig.height = 4}
for(i in 1:ncol(mapping.pheno)){
    plot(mean.scans, lodcol = i, map = map, main = colnames(mapping.pheno)[i])
}
```

The following heat map show the scans as above but
for mapping with strain means. Only LOD scores greater
than `r lod.thresh` are shown.

```{r multilod_strain, fig.width = 10, fig.height = 2.5, warning = FALSE}
multilod.plot(mean.scans, map = map, lod.thresh = lod.thresh, 
    row.name.shift = -3000, use.pheatmap.colors = TRUE, 
    border.col = "black", border.lwd = 1, legend.height = 6)
```


The following heat map shows all scans plotted together
so that the coordinates can be compared more easily.
All LOD scores above `r lod.thresh` are shown.

```{r all_scans, fig.width = 10, fig.height = 3}
colnames(mean.scans) <- paste(colnames(mean.scans), "mean", sep = "_")
colnames(pheno.scans) <- paste(colnames(pheno.scans), "ind", sep = "_")

all.scans <- cbind(pheno.scans, mean.scans)

multilod.plot(all.scans, map = map, 
    lod.thresh = lod.thresh, row.name.shift = -3500, use.pheatmap.colors = TRUE, 
    border.col = "black", border.lwd = 1, legend.height = 1)
```

The following heatmap shows the similarity between pairs of scans
for all the traits. The two primary groups are AR and Global/Classical.
Within each group, the mappings with individuals and with strain means
tend to map to the same place, which is reassuring. 

```{r scan_sim, fig.width = 7, fig.height = 7}
pheatmap(cor(pheno.scans))
```

The following heatmap shows all the scans again, this time 
with the traits presented in the same order as the clustered
correlation matrix above. This ordering better highlights where 
multiple scans are mapping to the same location. 

```{r clustered_heatmap, fig.width = 10, fig.height = 3}

trait.order <- hclust(dist(cor(all.scans)))$order

multilod.plot(all.scans[,trait.order], map = map, 
    lod.thresh = 2, row.name.shift = -3500, use.pheatmap.colors = TRUE, 
    border.col = "black", border.lwd = 1, legend.height = 1)
```

## Batch effects

The following code compares pairwise batches
to each other to look for systematic differences
between them

```{r batch_comparisons}
batch.pairs <- pair.matrix(1:length(u_batch))
batch.test <- vector(mode = "list", length = nrow(batch.pairs))
compared.strains <- rep(NA, nrow(batch.pairs))
for(bp in 1:nrow(batch.pairs)){
    batch1 <- batch.pairs[bp,1]
    batch2 <- batch.pairs[bp,2]
    cat("####", u_batch[batch1], "vs.", u_batch[batch2], "\n")
    shared.strain <- rownames(merged.batch)[intersect(which(merged.batch[,batch1] > 0), which(merged.batch[,batch2] > 0))]
    compared.strains[bp] <- paste(shared.strain, collapse = ", ")
    test <- cds.by.batch(shared.strain, batch.num = c(batch1, batch2), plot.type = "strip",
        plot.label = paste(u_batch[batch1], "vs.", u_batch[batch2]), remove.zeros = FALSE)

    batch.test[[bp]] <- test[[1]]

    cat("\n\n")
}
```


```{r merge_subsets}

all.subsets <- vector(mode = "list", length = length(u_batch))
for(b in 1:length(u_batch)){
    batch.idx <- which(batch.pairs == b, arr.ind = TRUE)
    subsets <- vector(mode = "list", length = nrow(batch.idx))
    for(su in 1:nrow(batch.idx)){
        subsets[[su]] <- batch.test[[batch.idx[su,1]]][[batch.idx[su,2]]]
    }
    names(subsets) <- paste("subset", 1:length(subsets))
    #plot.grouped.boxes(subsets, type = "matrix", main = u_batch[b])
    all.subsets[[b]] <- subsets
}

test <- lapply(all.subsets, function(x) Reduce("rbind", x))
names(test) <- u_batch
no.zero <- lapply(test, function(x) apply(x, 2, function(y) y[which(y > 0)]))
plot.grouped.boxes(test, type = "matrix", plot.type = "strip")
plot.grouped.boxes(no.zero, plot.type = "strip")
```

## LOD heat maps

It is easier to compare QTL heights and positions if 
we can put all the traces on a single plot. The 
following figures show multiple scans in one panel
so we can compare them more easily.

The following heatmap shows LOD scores across all 
traits mapped with individuals.

```{r lod_thresh}
lod.thresh <- 2
```

Only LOD scores greater than `r lod.thresh` are shown.

```{r, multilod_ind, fig.width = 10, fig.height = 2.5, warning = FALSE, message = FALSE}
multilod.plot(pheno.scans, map = map, lod.thresh = lod.thresh, row.name.shift = -2000, 
    use.pheatmap.colors = TRUE, border.col = "black", border.lwd = 1, legend.height = 6)
```



## Sex effects and strain means

For now I am removing the mapping using sex as an interactive 
covariate. Sex does not have a significant effect on these 
traits overall, and I'm concerned that a strain imbalance 
in a sex effect (random or otherwise) is inflating the LOD 
scores (see excised code).

I am also removing the mapping with strain means. The variance
within each strain is so high that I think mapping with strain 
means is misleading at best. 