---
title: "Prioritization"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
bibliography: eae.bib
---


## Introduction

This workflow attempts to prioritize genes in QTL identified
in Mapping_EAE_traits.Rmd


I downloaded GWAS results associated with Multiple Sclerosis
from the GWAS catalog on September 21, 2023. We will use
the genes associated with the most significant SNPs for 
our prioritization.

```{r setup}
rm(list = ls())
library(here)

args <- commandArgs(trailingOnly=T)
tissue.type = args[1]
lookup.genes <- as.logical(args[2]) #whether to do pubmed query at end

num.ms.genes <- 500

if(is.null(tissue.type) || is.na(tissue.type)){
    tissue.type = "central nervous system"; lookup.genes = FALSE
    #tissue.type = "hemolymphoid system"; lookup.genes = FALSE
    #tissue.type = "spleen"; lookup.genes = FALSE
}


#there is a lot of nesting in the directory structure, which is 
#for when we are running multiple modules and multiple traits, 
#and all that. It's excessive here, but it's the way the TRiAGE
#functions are written
project.name = "QTL_Prioritization"

results.dir <- here("Results", "Prioritization", paste0(gsub(" ", "_", tissue.type, "_network")))

if(!file.exists(results.dir)){dir.create(results.dir, recursive = TRUE)}
```

This workflow uses the `r tissue.type` network.

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE, pattern = ".R")
for(i in 1:length(all.fun)){source(all.fun[i])}
```


```{r load_libraries, warning = FALSE, error = FALSE, message = FALSE, echo = FALSE}
needed.libraries <- c("Matrix", "knitr", "here", "igraph", "DescTools", "pheatmap", 
"e1071", "parallel", "doParallel", "gprofiler2", "clusterProfiler", "DT", "easyPubMed",
"plotly", "qtl2", "cluster") 
load_libraries(needed.libraries)
n.cores <- detectCores() - 1
```

## GWAS

The following plot shows the distribution of -log10(p values)
for the genes in this list with a subset of the genes labeled.
We took the top `r num.ms.genes` for training and prioritization.

```{r gwas}
gwas <- read.delim(here("Data", "human", "gwas-association-downloaded_2023-09-21-EFO_0009585-withChildTraits.tsv"))
mapped.gene <- gwas[,"MAPPED_GENE"]
reported.gene <- gwas[,"REPORTED.GENE.S."]
pvals <- gwas[,"P.VALUE"]
gene.table <- cbind(mapped.gene, reported.gene, pvals)

#find all p values for each gene
split.mapped <- multi.strsplit(mapped.gene, c(", ", " - ", "; "))
split.reported <- multi.strsplit(reported.gene, c(", ", " - ", "; "))
u_genes <- unique(split.mapped, split.reported)

all.gene.p <- vector(mode = "list", length = length(u_genes))
names(all.gene.p) <- u_genes
for(g in 1:length(u_genes)){
    gene.p <- NULL
    mapped.idx <- grep(u_genes[g], mapped.gene)
    if(length(mapped.idx) > 0){
        gene.p <- c(gene.p, pvals[mapped.idx])
    }
    
    reported.idx <- grep(u_genes[g], reported.gene)
    if(length(reported.idx) > 0){
        gene.p <- c(gene.p, pvals[reported.idx])
    }
    all.gene.p[[g]] <- gene.p
}

neg.log.gene <- lapply(all.gene.p, function(x) -log10(x))
gene.order <- order(sapply(neg.log.gene, mean), decreasing = FALSE)

par(mar = c(4,6,4,4))
boxplot(neg.log.gene[gene.order], las = 2, horizontal = TRUE, cex.names = 0.5)

top.genes <- names(neg.log.gene)[gene.order[1:num.ms.genes]]
```

The top genes are enriched for the following processes.

```{r top_enrich}
enrich <- gost(top.genes, sources = c("GO", "HP", "KEGG", "REACTOME", "CORUM"))
plot.enrichment(enrich, order.by = "p_value")
```

After selecting training genes, we need to identify the mouse orthologs
and their Entrez IDs. Not all genes will have mouse orthologs

```{r convert}
mus.hum <- read.delim(here("Data", "mouse", "human.mouse.orthologs.txt"))
gene.idx <- match(top.genes, mus.hum[,"Human.Gene.Name"])
mus.entrez <- mus.hum[gene.idx,"Mouse.Ortholog.Entrez"]
mus.entrez <- mus.entrez[which(!is.na(mus.entrez))]

input.table <- mus.hum[gene.idx,]
```

## FNTM

```{r download_net, eval = TRUE, echo = FALSE}
#Download the network if it hasn't been previously downloaded. 
tissue.name = gsub(" ", "_", tissue.type)
net.file <- here("Data", "FGN", paste0(tissue.name, "_top.RData"))
if(!file.exists(net.file)){
    download.tissue.net(tissue.name, organism = "mouse", top.edges.only = TRUE, 
    project.dir = here("Data", "FGN"))
    }
```


```{r read_net, echo = FALSE}
#Read in the downloaded network.
all.var <- ls()
net.loaded <- as.logical(length(which(all.var == "tissue.net")))
if(!net.loaded){
    tissue.net <- readRDS(net.file)
}
```

Get information for all genes listed.

```{r gene_info}
gene.info <- read.delim(here("Data", "mouse", "mouse_gene_info.txt"))
```

### Identify Candidate Genes


First pull out genes from each candidate region 
and remove any from the training set.

```{r candidate_table_fun}

#only load package and library if we need to
#lib.loaded <- as.logical(length(which(all.var == "mus")))
#if(!lib.loaded){
#    library(biomaRt)
#    mus <- useEnsembl(biomart="ensembl", dataset="mmusculus_gene_ensembl", 
#    host = "www.ensembl.org")
    #mus <- useMart("ENSEMBL_MART_ENSEMBL", dataset = "mmusculus_gene_ensembl", 
    #host = "may2017.archive.ensembl.org")
#}


map <- readRDS(here("Data", "map.RDS"))

genes_in_region <- function(chr, pos.min, pos.max){
    chr.idx <- which(gene.info[,"chromosome_name"] == chr)
    above.min <- which(gene.info[,"start_position"] >= pos.min)
    below.max <- which(gene.info[,"start_position"] <= pos.max)
    gene.idx <- Reduce("intersect", list(chr.idx, above.min, below.max))
    region.table <- gene.info[gene.idx,]
    return(region.table)
}

qtl_candidate_table <- function(scan1_results, chr, lodthresh, prob = 0.95, Mb.buffer = 0){

    peak.table <- find_peaks(scan1_results, map = map, threshold = lodthresh, prob = prob)
    chr.idx <- which(peak.table[,"chr"] == chr)
    lod.int <- peak.table[chr.idx,c("ci_lo", "ci_hi")]
    candidate.table <- genes_in_region(chr, (as.numeric(lod.int[1])-Mb.buffer)*1e6, (as.numeric(lod.int[2])+Mb.buffer)*1e6)

    result <- list("candidate_table" = candidate.table, "interval" = lod.int)

    return(result)
}
```

```{r qtl_candidates}
test.qtl <- read.delim(here("Data", "qtl_data.txt")) #made by hand for ease of editing the QTLs of interest

qtl.results <- vector(mode = "list", length = nrow(test.qtl))
names(qtl.results) <- test.qtl[,1]

for(i in 1:nrow(test.qtl)){
    scan1.result <- readRDS(here("Results", "Zeros_included_covar-batch_rank-norm", test.qtl[i,"results_file"]))
    trait.idx <- which(names(scan1.result) == test.qtl[i,"qtl_name"])
    qtl_table <- qtl_candidate_table(scan1.result[[trait.idx]], 
        chr = test.qtl[i,"Chr"], lodthresh = test.qtl[i,"lodthresh"], prob = test.qtl[i,"prob"])
    qtl.results[[i]] <- qtl_table
}

all_candidates <- Reduce("rbind", lapply(qtl.results, function(x) x$candidate_table))
write.table(all_candidates, here("Results", "Prioritization", "qtl_candidates.csv"), 
    sep = ",", quote = FALSE, row.names = FALSE)
all_lod_int <- lapply(qtl.results, function(x) x$interval)
names(all_lod_int) = c("14", "4")

overlapping.genes <- intersect(all_candidates[,"entrezgene_id"], mus.entrez)
```

There is/are `r length(overlapping.genes)` positional candidates that
are also in the GWAS training set. We removed these from the training
set.

```{r}
overlap.idx <- match(overlapping.genes, all_candidates[,"entrezgene_id"])
overlapping.names <- all_candidates[overlap.idx,"external_gene_name"]
kable(all_candidates[overlap.idx,])

to.remove <- which(mus.entrez %in% overlapping.genes)
to.keep <- setdiff(1:length(mus.entrez), to.remove)
training.set <- mus.entrez
```
``

## Setup Positive-Labeled Gene Features

Get the connection weights of the positive-labeled genes to each other
in the tissue network. Not all of the genes in the list will have
Entrez IDs, and not all of the genes will have connections to the 
other genes in the tissue network. We trim off positive-labeled genes 
that are not connected to any other positive-labeled genes.


```{r pick_genes, echo = FALSE}
tissue.adj <- tissue.adj.mat(tissue.net, training.set, inc.all.genes = FALSE)
```

After identifying all Entrezgene IDs and building the positive-labeled network,
we have `r nrow(tissue.adj)` positive-labeled genes for training.

## Train SVM
We used this positive-labeled network as features to train 100 SVMs to classify
MS- and not MS-related genes. In each of the 100 SVMs, we selected a balanced
set of unlabeled genes for training. We write each of the trained models to a 
file for further analysis.

```{r train_models, message = FALSE, warning = FALSE, error = FALSE, eval = TRUE}
generate.triage.models(path = results.dir, project.name = project.name, 
trait.genes = training.set, fgn = tissue.net, n.trials = 100, cluster.modules = FALSE, 
cluster.threshold = 700, max.cluster.size = 600, min.cluster.size = 10, 
use.SVD = FALSE, verbose = TRUE, n.cores = n.cores)
```

## ROC Curves

We calculated ROC curves for each SVM, as well as the overal AUC for
the group. The plot below shows the ROC curves for all 100 models
along with the average AUC.

```{r roc, fig.height = 5, fig.width = 5}
all.fptp <- svm.ROC(results.dir, n.cores = n.cores, plot.results = TRUE)
all.auc <- sapply(all.fptp, function(x) smooth.auc(Reduce("rbind", x), plot = FALSE))
saveRDS(all.auc, file.path(results.dir, "AUC.RDS"))
```

## Score QTL genes

We then used the trained models to classify genes under the 
previously identified QTL.

Because of the way TRiAGE is set up, we will score all candidates
in all QTLs and then separate the results for viewing.


### Score genes

We scored all genes in the genome, which helps us identify
the functional clusters we are identifying that that are 
related to the GWAS hits used in the training set.

```{r score_genes}
score.candidates(results.dir, candidate.genes = gene.info[,"entrezgene_id"],
    verbose = TRUE, plot.results = FALSE)
merge.svm.gene.info(results.dir, gene.info)
```

The following plot shows the distribution of functional
scores (-log10(FPR)). The red line shows where the 99th
percentile cutoff is.

```{r score_dist}
score.table <- read.csv(file.path(results.dir, "QTL_Prioritization", "Module1", "Candidate.Gene.Results.csv"))
fun.score <- -log10(score.table[,"Mean.FP.Rate"])
names(fun.score) <- score.table[,"entrezgene_id"]
hist(fun.score, main = "Functional Score", xlab = "Functional Score", breaks = 100)

top.perc = 99
top.score <- get.percentile(fun.score, top.perc)
abline(v = top.score, col = "red")
n.top <- length(which(fun.score >= top.score))
```

We looked at the adjacency matrix of the top `r n.top` genes,
which represents the top `r top.perc` percentile of functional
scores. 

The network of these top scoring genes doesn't have a huge 
amount of structure. Any clusters in it are also highly related 
to each other.

```{r top_net}
score.order <- order(fun.score, decreasing = TRUE)
top.genes <- names(fun.score[score.order])[1:n.top]
#plot(fun.score[score.order][1:n.top])

#create a new network with the top genes from the scoring
top.adj <- tissue.adj.mat(tissue.net, top.genes, inc.all.genes = FALSE)
#pheatmap(top.adj, show_rownames = FALSE, show_colnames = FALSE)
```

```{r top_cl}
#look for clusters
#net.decomp <- plot.decomp(top.adj, pc = 4)
test <- test.pam.k(top.adj, kseq = 2:10, diss = FALSE, metric = "euclidean", plot.results = FALSE)
#boxplot(test$cl.width)
num.cl <- names(which.max(sapply(test$cl.width, mean)))
```

We used a silhouette method to identify the best number
of clusters in the matrix using k means clustering. This
was `r num.cl` clusters, shown below.

```{r cl_def, fig.width = 8, fig.height = 7}
cl.mem <- test$mem[,num.cl]
cl.df <- data.frame("cluster" = as.factor(cl.mem))
cl.order <- order(cl.mem)
#pheatmap(top.adj[cl.order,cl.order], show_rownames = FALSE, show_colnames = FALSE)

#pheatmap(top.adj[cl.order,cl.order])
#cl.tree <- hclust(dist(top.adj))
#cl.mem <- cutree(cl.tree, k = 2)
cl.df <- data.frame("cluster" = as.factor(cl.mem))
cl.order <- order(cl.mem)
pheatmap(top.adj[cl.order,cl.order], cluster_rows = FALSE, cluster_cols = FALSE,
    show_rownames = FALSE, show_colnames = FALSE, annotation_row = cl.df)
```

The functional enrichments of these clusters are shown below.
These are the functions that are most highly related to the 
input training set of genes according to the `r tissue.type`
network.

```{r cl_enrich, fig.width = 7, fig.height = 7}

u_cl <- unique(cl.mem)
cl.id <- lapply(u_cl, function(x) names(cl.mem)[which(cl.mem == x)])
cl.enrich <- lapply(cl.id, function(x) gost(x, organism = "mmusculus", 
    sources = c("GO", "KEGG", "REACTOME"), numeric_ns = "ENTREZGENE_ACC"))
names(cl.enrich) <- paste0("Cluster", 1:length(cl.enrich))
plot.enrichment.group(cl.enrich, sort.by = "p_value", max.term.size = 3000, n.terms = 20)
```

### Rank Genes in Each QTL

We then looked more specifically at the genes in the QTL
identified in Mapping_EAE_traits.Rmd.

```{r get_fp, error = FALSE, warning = FALSE, message = FALSE}
fp.mat <- get.all.fp(results.dir)
#replace zeros with a minimum score
#so we don't end up with infinite scores
min.non.zero <- min(fp.mat[which(fp.mat[,1] > 0),1])
min.val <- min.non.zero/10
zero.locale <- which(fp.mat == 0)
fp.mat[zero.locale,1] <- min.val

neg.log.fp <- -log10(fp.mat)

common.genes <- intersect(rownames(neg.log.fp), all_candidates[,"external_gene_name"])
result.table <- cbind(all_candidates[match(common.genes, all_candidates[,"external_gene_name"]),], neg.log.fp[common.genes,])
colnames(result.table)[ncol(result.table)] <- "neg.log.fpr"
```

## Gene Scores by Position

The following plots show the genes plotted in their genomic
position on the x axis and their -log(FPR) on the y axis.
The higher the gene on the y axis the more confident the 
model is that it is related to the input genes. Vertical 
gray dashed lines show the boundaries of the QTL. The
horizontal lines correspond to an FPR of 0.05 and 0.1.


```{r get_region}

u_chr <- unique(result.table[,"chromosome_name"])

for(ch in 1:length(u_chr)){
    chr.idx <- which(result.table[,"chromosome_name"] == u_chr[ch])
    gene.pos <- result.table[chr.idx,"start_position"]/1e6
    pos.region <- max(gene.pos) - min(gene.pos)
    xmin <- min(gene.pos)-(pos.region*0.05)
    xmax <- max(gene.pos)+(pos.region*0.05)
    yvals <- result.table[chr.idx,"neg.log.fpr"]
    ymax <- max(yvals)*1.05

    plot.col = rep("black", length(gene.pos))
    has.gwas <- which(result.table[chr.idx,"external_gene_name"] %in% overlapping.names)
    if(length(has.gwas) > 0){
        plot.col[has.gwas] <- "red"
    }

    plot(gene.pos, yvals, ylim = c(0, ymax),
        xlab = paste("Chr", u_chr[ch], "Position (Mb)"), ylab = "-log10(FPR)",
        main = paste("Chr", u_chr[ch], "QTL"), pch = 16, xlim = c(xmin, xmax))
    text(result.table[chr.idx,"start_position"]/1e6, result.table[chr.idx,"neg.log.fpr"],
        labels = result.table[chr.idx, "external_gene_name"], pos = 3)
    int.idx <- which(names(all_lod_int) == u_chr[ch])
    abline(v = all_lod_int[[int.idx]], lty = 2, col = "gray")
    abline(h = -log10(c(0.05, 0.1)), lty = c(1,2))

    #plot unranked genes along the x axis. Mark any that 
    #are GWAS hits.
    ranked.genes <- result.table[chr.idx, "external_gene_name"]
    all.region.genes <- genes_in_region(chr = u_chr[ch], 
        min(result.table[chr.idx,"start_position"]), 
        max(result.table[chr.idx,"start_position"]))
    unranked.genes <- setdiff(all.region.genes[,"external_gene_name"], ranked.genes)
    is.gwas <- match(unranked.genes)



}
```


## Ranked Genes by QTL

The following plots show only the scores of each gene
independent of position. They are sorted from highest
to lowest score. The black vertical lines correspond to
FPR of 0.05 and 0.1.

The histograms show the distribution of functional scores
across the mouse genome and where the top-ranked gene in 
the locus fell in that distribution.

```{r, fig.width = 8, fig.height = 4}
for(ch in 1:length(u_chr)){
    par(mfrow = c(1,2))
    chr.idx <- which(result.table[,"chromosome_name"] == u_chr[ch])
    result.fpr <- result.table[chr.idx,"neg.log.fpr"]
    result.order <- order(result.fpr, decreasing = FALSE)
    par(mar = c(4,8,4,2))
    barplot(result.fpr[result.order], las = 2, 
        main = paste("Chr", u_chr[ch], "QTL"),
        ylab = "", names = result.table[chr.idx[result.order],"external_gene_name"],
        horiz = TRUE)
    abline(v = -log10(c(0.05, 0.1)), lty = c(1,2))
    mtext("-log10(FPR)", side = 2, line = 4.5)

    par(mar = c(4,4,4,4))
    hist(fun.score, xlab = "Functional Score", main = "All Functional Scores")
    abline(v = max(result.table[chr.idx,"neg.log.fpr"]), col = "red")
    legend("topright", lty = 1, col = "red",
        legend = tail(result.table[chr.idx[result.order],"external_gene_name"], 1))
}

```

```{r pubmed, results = "asis"}
if(lookup.genes){
    cat("## Literature associations\n")
    all_genes <- result.table[,"external_gene_name"]
    all.pubmed <- lapply(all_genes, function(x) get_pubmed_ids(paste("Multiple Sclerosis AND", x)))

    num.papers <- unlist(lapply(all.pubmed, function(x) x$Count))
    with.papers <- which(num.papers != 0)
    paper.counts <- cbind(result.table[with.papers,"external_gene_name"], num.papers[with.papers])
    number.order <- order(as.numeric(paper.counts[,2]))
    barplot_with_num(as.numeric(paper.counts[number.order,2]), 
        names = paper.counts[number.order,1], las = 2, ylab = "Number of Abstracts")


    has.data <- all.pubmed[with.papers]
    all.abstracts <- lapply(has.data, function(x) fetch_pubmed_data(x, format = "abstract"))
    for(i in 1:length(all.abstracts)){
        write.table(all.abstracts[[i]], here("Results", "Prioritization", tissue.name, 
            paste0("PubMed_Query_", all_genes[with.papers[i]], ".txt")), 
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
    }

}
```